---
title: "Assignment 4"
subtitle: "Parking congestion in Melbourne CBD"
author: "Vishal Sampath Kumar"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
editor: visual
---

### Data Details

#### Topic of this work

Which Melbourne parking zones are most consistently congested, and how does that congestion vary by time of day and day of week?

#### Why this topic

I chose this topic because I live near the CBD, whenever I have friends coming over or we plan a night out, figuring out parking is always a painpoint and an item of discussion that has caused many plans to fall through before, which is what led me to investigate this topic and shape my research question around parking congestion. This topic is engaging to me because it connects everyday urban movement data with practical questions about how city space is used, which makes the project practical and interesting.

#### Data description

I used the datasets described below to analyse recurring parking congestion in Melbourne. Together, these files provide timestamped bay status information and spatial location data, which makes them suitable for both time-based and map-based analysis.

**Datasets:**

1.  **On-street Parking Bay Sensors**

    - Source: City of Melbourne Open Data Portal.
    - Link: (data.melbourne.vic.gov.au). [On-street Parking Bay Sensors](https://data.melbourne.vic.gov.au/explore/dataset/on-street-parking-bay-sensors/information/?dataChart=eyJxdWVyaWVzIjpbeyJjaGFydHMiOlt7InR5cGUiOiJjb2x1bW4iLCJmdW5jIjoiQ09VTlQiLCJ5QXhpcyI6ImJheV9pZCIsInNjaWVudGlmaWNEaXNwbGF5Ijp0cnVlLCJjb2xvciI6InJhbmdlLWN1c3RvbSJ9XSwieEF4aXMiOiJsYXN0dXBkYXRlZCIsIm1heHBvaW50cyI6NTAsInNvcnQiOiIiLCJ0aW1lc2NhbGUiOiJkYXkiLCJjb25maWciOnsiZGF0YXNldCI6Im9uLXN0cmVldC1wYXJraW5nLWJheS1zZW5zb3JzIiwib3B0aW9ucyI6e319LCJzZXJpZXNCcmVha2Rvd24iOiJ6b25lX251bWJlciJ9XSwidGltZXNjYWxlIjoiIiwiZGlzcGxheUxlZ2VuZCI6dHJ1ZSwiYWxpZ25Nb250aCI6dHJ1ZX0%3D) .
    - Content: Live parking bay sensor data showing whether a vehicle is present or not present, along with spatial coordinates and kerbside ID.
    - Data type: Observational administrative open data, effectively a live operational record rather than a sample or experiment.
    - Update frequency: Approximately every two minutes.
    - Licence: CC BY / Creative Commons Attribution 4.0.
    - Limitations: Depends on sensor reliability, may contain missing values, and reflects bay status rather than the full context of driver behaviour.
    - Privacy/Ethics: Low privacy risk because the data concerns parking bays rather than named individuals, but the analysis will remain aggregated and will not attempt to identify vehicles or people.

2.  **On-street Parking Bay Sensors Spatial File (GeoJSON or SHP)**

    - Source: Same City of Melbourne dataset page.
    - Link: (data.melbourne.vic.gov.au). [On-street Parking Bay Sensors](https://data.melbourne.vic.gov.au/explore/dataset/on-street-parking-bay-sensors/information/?dataChart=eyJxdWVyaWVzIjpbeyJjaGFydHMiOlt7InR5cGUiOiJjb2x1bW4iLCJmdW5jIjoiQ09VTlQiLCJ5QXhpcyI6ImJheV9pZCIsInNjaWVudGlmaWNEaXNwbGF5Ijp0cnVlLCJjb2xvciI6InJhbmdlLWN1c3RvbSJ9XSwieEF4aXMiOiJsYXN0dXBkYXRlZCIsIm1heHBvaW50cyI6NTAsInNvcnQiOiIiLCJ0aW1lc2NhbGUiOiJkYXkiLCJjb25maWciOnsiZGF0YXNldCI6Im9uLXN0cmVldC1wYXJraW5nLWJheS1zZW5zb3JzIiwib3B0aW9ucyI6e319LCJzZXJpZXNCcmVha2Rvd24iOiJ6b25lX251bWJlciJ9XSwidGltZXNjYWxlIjoiIiwiZGlzcGxheUxlZ2VuZCI6dHJ1ZSwiYWxpZ25Nb250aCI6dHJ1ZX0%3D) .
    - Content: Spatial representation of the parking sensor data that can be used for mapping sensor locations.
    - Data type: Spatial open data.
    - Licence: CC BY / Creative Commons Attribution 4.0.
    - Limitations: Spatial points help show location, but they do not on their own explain why some areas are more congested than others.
    - Privacy/Ethics: Public infrastructure data with low privacy risk.

3.  **On-street Parking Bays**

    - Source: City of Melbourne Open Data Portal.
    - Link: (data.melbourne.vic.gov.au). [On-street Parking Bay Sensors](https://data.melbourne.vic.gov.au/explore/dataset/on-street-parking-bay-sensors/information/?dataChart=eyJxdWVyaWVzIjpbeyJjaGFydHMiOlt7InR5cGUiOiJjb2x1bW4iLCJmdW5jIjoiQ09VTlQiLCJ5QXhpcyI6ImJheV9pZCIsInNjaWVudGlmaWNEaXNwbGF5Ijp0cnVlLCJjb2xvciI6InJhbmdlLWN1c3RvbSJ9XSwieEF4aXMiOiJsYXN0dXBkYXRlZCIsIm1heHBvaW50cyI6NTAsInNvcnQiOiIiLCJ0aW1lc2NhbGUiOiJkYXkiLCJjb25maWciOnsiZGF0YXNldCI6Im9uLXN0cmVldC1wYXJraW5nLWJheS1zZW5zb3JzIiwib3B0aW9ucyI6e319LCJzZXJpZXNCcmVha2Rvd24iOiJ6b25lX251bWJlciJ9XSwidGltZXNjYWxlIjoiIiwiZGlzcGxheUxlZ2VuZCI6dHJ1ZSwiYWxpZ25Nb250aCI6dHJ1ZX0%3D) .
    - Content: Spatial polygons representing parking bays across the city.
    - Data type: Spatial administrative open data.
    - Licence/conditions: Public open data with City of Melbourne disclaimer noted on the dataset page.
    - Use in project: Optional supporting dataset for richer mapping and for understanding joins between the live parking release datasets.
    - Limitation: Not every related parking dataset joins directly to every other dataset; the sensor data joins to the parking bays data by `marker_id`.

4.  **Municipal Boundary**

    - Source: City of Melbourne / Victorian Government Data Directory.
    - Link: \[Municipal boundary\](https://data.melbourne.vic.gov.au/explore/dataset/municipal-boundary/.
    - Content: Boundary of the City of Melbourne municipality used as the basemap polygon for the spatial plots.
    - Data type: Spatial boundary data.
    - Use in project: Provides a recognisable Melbourne map so the parking points can be plotted on the municipality rather than on blank coordinate axes.

#### Data download

The following section describes how the data can be downloaded from its original sources and the initial processing steps used for analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(here)
library(lubridate)
library(sf)
library(janitor)
library(openxlsx)
library(viridis)
```

```{r}
dir.create(here::here("raw_data"), showWarnings = FALSE)
dir.create(here::here("data"), showWarnings = FALSE)
```

**Downloading Dataset 1**

```{r}
url_csv <- "https://data.melbourne.vic.gov.au/explore/dataset/on-street-parking-bay-sensors/download/?format=csv&timezone=Australia/Melbourne&lang=en"

destfile_csv <- here::here("raw_data/on-street-parking-bay-sensors.csv")
download.file(url_csv, destfile_csv, mode = "wb")

parking_sensor_raw <- read_delim(
  destfile_csv,
  delim = ";",
  show_col_types = FALSE
) |>
  clean_names()
```

If the direct download link changes, the file can be downloaded manually from the dataset page and saved as `raw_data/on-street-parking-bay-sensors.csv`.

**Downloading Dataset 2**

```{r}
url_geojson <- "https://data.melbourne.vic.gov.au/explore/dataset/on-street-parking-bay-sensors/download/?format=geojson&timezone=Australia/Melbourne&lang=en"

destfile_geojson <- here::here("raw_data/on-street-parking-bay-sensors.geojson")
download.file(url_geojson, destfile_geojson, mode = "wb")

parking_sensor_geo <- st_read(destfile_geojson, quiet = TRUE)
```

If working with the downloaded shapefile instead, all associated files (`.shp`, `.shx`, `.dbf`, `.prj`, `.cpg`) should be stored together in the same folder before reading the `.shp` file.

```{r}
# If using a shapefile instead of geojson:
# parking_sensor_geo <- st_read(here::here("raw_data/on-street-parking-bay-sensors/on-street-parking-bay-sensors.shp"))
```

**Downloading Dataset 3**

```{r}
url_bays <- "https://data.melbourne.vic.gov.au/explore/dataset/on-street-parking-bays/download/?format=csv&timezone=Australia/Melbourne&lang=en"

destfile_bays <- here::here("raw_data/on-street-parking-bays.csv")
download.file(url_bays, destfile_bays, mode = "wb")

parking_bays_raw <- read_delim(
  destfile_bays,
  delim = ";",
  show_col_types = FALSE
) |>
  clean_names()
```

This supporting dataset is useful for documenting how the City of Melbourne’s live parking datasets relate to each other.

#### Data cleaning and preparation

Before analysing congestion, the raw parking sensor data needs to be converted into a form that is easier to summarise. That includes parsing timestamps properly, extracting useful time-based variables such as hour and day of week, converting latitude and longitude into numeric form, and creating a simple logical indicator that marks whether a bay is occupied.

The main steps used to process the data are listed below.

1.  Import the raw CSV sensor data.
2.  Import the geojson or shapefile spatial data.
3.  Clean variable names.
4.  Parse timestamp fields into proper datetime format.
5.  Create derived variables such as hour, weekday, and date.
6.  Recode `status_description` into an occupied indicator.
7.  Remove or flag missing zone values.
8.  Split the coordinate field into latitude and longitude.
9.  Summarise occupancy by zone, hour, and weekday.
10. Save the cleaned analysis file separately from the raw data.

```{r}
parking_clean <- parking_sensor_raw |>
  mutate(
    lastupdated = ymd_hms(lastupdated),
    status_timestamp = ymd_hms(status_timestamp),
    hour = hour(status_timestamp),
    day_of_week = wday(status_timestamp, label = TRUE, week_start = 1),
    date = as_date(status_timestamp),
    occupied = case_when(
      str_trim(str_to_lower(status_description)) == "present" ~ TRUE,
      str_trim(str_to_lower(status_description)) == "unoccupied" ~ FALSE,
      TRUE ~ NA
    )
  ) |>
  separate(location, into = c("latitude", "longitude"), sep = ",", remove = FALSE) |>
  mutate(
    latitude = as.numeric(str_trim(latitude)),
    longitude = as.numeric(str_trim(longitude))
  ) |>
  filter(
    !is.na(zone_number),
    !is.na(status_timestamp),
    !is.na(hour),
    !is.na(day_of_week),
    !is.na(latitude),
    !is.na(longitude),
    !is.na(occupied)
  )

parking_clean

write_csv(parking_clean, here::here("data/parking_processed.csv"))
```

The cleaning step also removes rows with missing values in the key analytical fields. Based on the output shown below, the cleaned dataset contains `r nrow(parking_clean)` rows and `r ncol(parking_clean)` columns, which is the dataset used in the later summaries and visualisations

Since spatial summaries are needed, the processed summaries can later be joined to the spatial data. The City of Melbourne states that the live sensor dataset joins to the parking bays dataset by `marker_id`, and the sensor dataset joins to the parking restrictions dataset by `bay_id`.

#### Creating the data dictionary file

```{r}
data_dictionary <- tibble(
  variable = c(
    "lastupdated",
    "status_timestamp",
    "zone_number",
    "status_description",
    "kerbsideid",
    "location",
    "hour",
    "day_of_week",
    "date",
    "occupied"
  ),
  description = c(
    "Time the record was last updated",
    "Timestamp of the parking sensor reading",
    "Parking zone identifier",
    "Description of whether a car is present or not present",
    "Kerbside identifier",
    "Coordinate/location field",
    "Hour extracted from status_timestamp",
    "Weekday extracted from status_timestamp",
    "Calendar date extracted from status_timestamp",
    "Logical occupied/vacant indicator created for analysis"
  ),
  role_in_analysis = c(
    "Checks update recency",
    "Main time variable",
    "Main grouping variable",
    "Defines congestion proxy",
    "Optional finer location grouping",
    "Supports mapping",
    "Used for hourly summaries",
    "Used for weekly summaries",
    "Used for date-based summaries",
    "Main congestion measure"
  )
)

write.xlsx(data_dictionary, here::here("data/data-dictionary.xlsx"))
```

#### Files included in submission

The submission will include:

- the raw CSV file;
- the raw GeoJSON or shapefile;
- the optional raw parking bays file;
- the Melbourne municipal boundary GeoJSON;
- the processed CSV file;
- a readme file describing the source and folder structure;
- and a data dictionary saved as an `.xlsx` file.

This documentation makes it clear what was downloaded, what was changed, and what was used in the analysis.

## Blog post

### Why take a look at parking congestion data?

Parking in the Melbourne CBD is a familiar source of frustration. It affects everyday social plans, visitor access, and how long people spend circling for a spot. Because the City of Melbourne publishes live data from thousands of parking sensors, it is possible to move beyond anecdotal complaints and examine actual patterns of parking pressure.

This blog post asks three connected questions: First, **which zones appear most consistently occupied?** Second, **how does occupancy vary by hour of the day?** Third, **does occupancy also vary across the days of the week?**\
The emphasis is on repeated patterns rather than a one-off count, because a zone that stays busy across many windows is more important than one that only spikes briefly. By combining grouped summaries and maps, the analysis aims to identify not just where parking is full, but where it appears persistently full over time.

### Data Sources for Analysing Parking Occupancy in Melbourne

Data for visualising parking occupancy in Melbourne is available through the City of Melbourne’s public open data portal.

The main dataset used here is the parking sensor dataset, which records whether a bay is occupied or unoccupied at specific timestamps. This makes it suitable for constructing a simple measure of parking pressure based on observed occupancy.

The spatial file provides the coordinates needed for mapping, and the parking bays file serves as a supporting source for interpretation. In this analysis, congestion is treated as repeated occupancy rather than, for example, turnover or parking duration. That is an important limitation, but it still provides a clear and interpretable starting point.

To make the analysis defensible, congestion needs to be defined explicitly. Here, congestion is measured as repeated occupied readings, and “consistent congestion” is identified by looking for zones that remain occupied across many hour-by-zone windows rather than only one peak period. That makes the analysis more informative than a simple busiest-zone ranking. A parking zone is treated as more congested when its bays appear occupied more often, and especially when that elevated occupancy persists across multiple time periods rather than only appearing in a single spike.

### Inspecting the data

```{r}
glimpse(parking_clean)
glimpse(parking_sensor_geo)
glimpse(parking_bays_raw)
```

The cleaned dataset contains `r nrow(parking_clean)` observations and `r ncol(parking_clean)` variables, which indicates that the filtering and recoding steps have left a sizeable working dataset for analysis. The preview also shows the key variables needed later, including timestamps, zone identifiers, occupancy labels, and coordinate values

### What does the data tell us?

The City of Melbourne notes that the sensor dataset records whether a vehicle is present or not present and includes spatial coordinate information and kerbside ID. It also states that the data is updated frequently and can be affected by outages or delays, so the `lastupdated` field is useful for checking freshness.

The first useful summary is the zone-level occupancy table shown below.

#### Aspect 1: Zone patterns

```{r}
zone_summary <- parking_clean |>
  group_by(zone_number) |>
  summarise(
    records = n(),
    occ_rate = mean(occupied, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(occ_rate), desc(records))

zone_summary |>
  slice_head(n = 10)

```

Looking at the zone summary, the top ten zones all have an occupancy rate of 1. This means every recorded observation for those zones in the cleaned sample was classified as occupied. Among these, zone **7594** stands out with **16 records**, followed by zone **7525** with **11 records** and zone **7220** with **10 records**. Other highly occupied zones, such as 7089 and 7156, also show complete occupancy but with fewer observations.

This result suggests that some zones were effectively always full during the observed sample. At the same time, the `records` column matters for interpretation. A zone with perfect occupancy across many observations is stronger evidence of persistent congestion than a zone with perfect occupancy based on only a handful of rows. This means the most convincing evidence comes from zones that combine a very high occupancy rate with a reasonable number of records.

In general though, all these zones have relatively small record counts, which means their perfect occupancy rates should be interpreted with some caution but this is still an informative result. It suggests that **some zones in the dataset are consistently busy and may represent areas of especially high parking pressure**. However, a high mean occupancy on its own does not distinguish between zones that are always busy and zones that only appear in a limited subset of time periods. To improve on that, it is useful to look at persistence across hourly windows.

A second summary considers persistence across hourly windows rather than just the overall average.

```{r}
zone_hour <- parking_clean |>
  group_by(zone_number, hour) |>
  summarise(
    hour_occ = mean(occupied, na.rm = TRUE),
    .groups = "drop"
  )

threshold <- median(zone_hour$hour_occ, na.rm = TRUE)

persistent_zones <- zone_hour |>
  mutate(high = hour_occ >= threshold) |>
  group_by(zone_number) |>
  summarise(
    persist_share = mean(high, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(persist_share))

persistent_zones |>
  slice_head(n = 10)

```

The persistence table shows that the top ten zones all have a `persist_share` of 1. In this case, that means these zones were above the median hourly occupancy threshold in every observed hourly period used in the calculation. The top group includes zones **7002**, **7015**, **7019**, **7053**, **7089**, **7096**, **7156**, **7160**, **7182**, and **7185**.

This is important because it shows that some zones are not merely busy on average, but repeatedly busy across time. The overlap between this persistence table and the earlier occupancy ranking, especially for zones like **7089** and **7156**, suggests that a smaller subset of zones experiences more durable congestion than the rest of the network.

These summaries make it possible to distinguish between zones that are always busy and zones that are only occasionally busy. A zone with a high `occ_rate` and a high `persist_share` is a stronger candidate for “consistent congestion” than a zone that only has a high reading in one or two periods.

#### Aspect 2: How does occupancy change through the day?

The hourly summary gives a clearer picture of when parking demand is highest.

```{r}
hour_summary <- parking_clean |>
  group_by(hour) |>
  summarise(
    occ_rate = mean(occupied, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(hour)

hour_summary

```

```{r}
ggplot(hour_summary, aes(x = hour, y = occ_rate)) +
  geom_col(fill = "#2a6fdb") +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  labs(
    title = "Occupancy rate by hour",
    x = "Hour of day",
    y = "Occupancy rate"
  ) +
  theme_minimal()

```

The hourly pattern shows when parking demand is strongest across the day. This helps answer whether congestion is tied to business hours, evening activity, or a broader all-day pattern.

The hourly pattern is clearly uneven. **Occupancy is relatively high overnight and in the early hours of the morning**, with **hour 0** reaching roughly **0.69**. It remains moderate through the early morning, then gradually declines through the late morning and early afternoon.

**The lowest point appears around hour 13**, where the occupancy rate falls to about **0.10**. After that, **occupancy begins to rise again through the late afternoon**. The strongest peak occurs around **hour 18**, where the occupancy rate reaches roughly **0.77**, before **remaining elevated into the evening**.

There is also a relatively high level of occupancy at **23:00** and around **midnight**, which suggests that busy periods are **not limited to standard business hours**.

This suggests that, in this sample, **the most difficult time to find parking is likely to be late afternoon and evening** rather than the middle of the day. That is a useful finding because it suggests the pressure on parking may be linked not only to daytime work activity, but also to later city activity.

#### Aspect 3: How does occupancy change by day of week

The next step is to check whether occupancy patterns also differ by weekday. Weekday patterns help distinguish work-related parking demand from weekend or leisure-related demand. If the busiest days differ from the busiest hours, that suggests parking congestion is shaped by more than one urban activity pattern.

```{r}
weekday_summary <- parking_clean |>
  group_by(day_of_week) |>
  summarise(
    occ_rate = mean(occupied, na.rm = TRUE),
    .groups = "drop"
  )

weekday_summary
```

The weekday summary shows a clear difference between weekdays and weekends. **Monday has the highest observed occupancy rate at about 0.589**, followed by Wednesday and Friday, while **Saturday is the lowest at about 0.354** and Sunday remains below 0.4

```{r}
ggplot(weekday_summary, aes(x = day_of_week, y = occ_rate)) +
  geom_col(fill = "#d95f02") +
  labs(
    title = "Occupancy rate by day of week",
    x = "Day of week",
    y = "Occupancy rate"
  ) +
  theme_minimal()
```

The bar chart reinforces that result visually. **Weekday occupancy is consistently higher than weekend occupancy**, with the sharpest drop occurring on Saturday, and only a slight recovery on Sunday. **Thursday is also somewhat lower than the other weekdays**, but still noticeably above the weekend values.

This suggests that the **strongest parking pressure in the sample is associated more with weekday urban activity** than with weekend demand. In other words, the busiest days are not evenly spread across the week.

**V1 Reflection:**

In this early version, I focused only on getting the data loaded into R. Instead of manually downloading the dtasets from the City of Melbourne website as I did on the submission for early feedback, and then loading them into R, this time I tried to download data directly from the url on Rstudio. However, this time, I noticed that when I download the datasets from the url using R markdown, the downloaded CSV files used a semicolon delimiter, so the dataset loaded as a single column. This is something I will have to fix before proceeding with any further analysis, so at this stage, the main focus is simply getting the file structure and dataset loading working.

**V2 Reflection:**

In this version, after realizing the delimiter problem, I fixed the import issue from the first draft using `read_delim(..., delim = ";")` and successfully imported the data into a format that could actually be analysed instead of being read as one long column. I then began cleaning the key variables so the timestamps, zone numbers, and occupancy values could be used properly and also started creating the time-based fields like hour and day of week.

Another thing I learned at this stage was that the location field needed extra wrangling before it could be used in maps. Splitting coordinates into latitude and longitude and checking for missing values took a bit of trial and error, but it was necessary before I could move on to any spatial visualisation. This version still feels very preliminary, but is a major step forward because the dataset is now structured enough to support summaries by time and place and I've also added a data dictionary file to supplement understanding.

**V3 Reflection:**

In this version, I moved beyond just cleaning the data and started building the actual analysis, commentating it along the way with my blog post. The biggest change from the earlier version was that I could now summarise occupancy by zone and time, which meant I was finally able to answer the research question in a more direct way. Additionally, I added more details about the sources and file structure in the README file and also decided to add a spatial aspect to my project by incorporating the municipal boundary and plotting the parking observations on a map, but haven't implemented it yet.

This stage also showed me that I needed a clearer way of defining “consistent congestion,” because just looking at the busiest zones was not enough. That led me to start thinking more carefully about persistence across time, rather than only average occupancy. At this point, the project is still missing some refinement, but it has already become much stronger because the analysis now combined time-based summaries, which gave me a better sense of what the final report would look like.
