---
title: "Assignment 4"
subtitle: "Parking congestion in Melbourne CBD"
author: "Vishal Sampath Kumar"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
editor: visual
---

### Data Details

#### Topic of this work

Which Melbourne parking zones are most consistently congested, and how does that congestion vary by time of day and day of week?

#### Why this topic

I chose this topic because I live near the CBD, whenever I have friends coming over or we plan a night out, figuring out parking is always a painpoint and an item of discussion that has caused many plans to fall through before, which is what led me to investigate this topic and shape my research question around parking congestion. This topic is engaging to me because it connects everyday urban movement data with practical questions about how city space is used, which makes the project practical and interesting.

#### Data description

I used the datasets described below to analyse recurring parking congestion in Melbourne. Together, these files provide timestamped bay status information and spatial location data, which makes them suitable for both time-based and map-based analysis.

**Datasets:**

1.  **On-street Parking Bay Sensors**

    - Source: City of Melbourne Open Data Portal.
    - Link: (data.melbourne.vic.gov.au). [On-street Parking Bay Sensors](https://data.melbourne.vic.gov.au/explore/dataset/on-street-parking-bay-sensors/information/?dataChart=eyJxdWVyaWVzIjpbeyJjaGFydHMiOlt7InR5cGUiOiJjb2x1bW4iLCJmdW5jIjoiQ09VTlQiLCJ5QXhpcyI6ImJheV9pZCIsInNjaWVudGlmaWNEaXNwbGF5Ijp0cnVlLCJjb2xvciI6InJhbmdlLWN1c3RvbSJ9XSwieEF4aXMiOiJsYXN0dXBkYXRlZCIsIm1heHBvaW50cyI6NTAsInNvcnQiOiIiLCJ0aW1lc2NhbGUiOiJkYXkiLCJjb25maWciOnsiZGF0YXNldCI6Im9uLXN0cmVldC1wYXJraW5nLWJheS1zZW5zb3JzIiwib3B0aW9ucyI6e319LCJzZXJpZXNCcmVha2Rvd24iOiJ6b25lX251bWJlciJ9XSwidGltZXNjYWxlIjoiIiwiZGlzcGxheUxlZ2VuZCI6dHJ1ZSwiYWxpZ25Nb250aCI6dHJ1ZX0%3D) .
    - Content: Live parking bay sensor data showing whether a vehicle is present or not present, along with spatial coordinates and kerbside ID.
    - Data type: Observational administrative open data, effectively a live operational record rather than a sample or experiment.
    - Update frequency: Approximately every two minutes.
    - Licence: CC BY / Creative Commons Attribution 4.0.
    - Limitations: Depends on sensor reliability, may contain missing values, and reflects bay status rather than the full context of driver behaviour.
    - Privacy/Ethics: Low privacy risk because the data concerns parking bays rather than named individuals, but the analysis will remain aggregated and will not attempt to identify vehicles or people.

2.  **On-street Parking Bay Sensors Spatial File (GeoJSON or SHP)**

    - Source: Same City of Melbourne dataset page.
    - Link: (data.melbourne.vic.gov.au). [On-street Parking Bay Sensors](https://data.melbourne.vic.gov.au/explore/dataset/on-street-parking-bay-sensors/information/?dataChart=eyJxdWVyaWVzIjpbeyJjaGFydHMiOlt7InR5cGUiOiJjb2x1bW4iLCJmdW5jIjoiQ09VTlQiLCJ5QXhpcyI6ImJheV9pZCIsInNjaWVudGlmaWNEaXNwbGF5Ijp0cnVlLCJjb2xvciI6InJhbmdlLWN1c3RvbSJ9XSwieEF4aXMiOiJsYXN0dXBkYXRlZCIsIm1heHBvaW50cyI6NTAsInNvcnQiOiIiLCJ0aW1lc2NhbGUiOiJkYXkiLCJjb25maWciOnsiZGF0YXNldCI6Im9uLXN0cmVldC1wYXJraW5nLWJheS1zZW5zb3JzIiwib3B0aW9ucyI6e319LCJzZXJpZXNCcmVha2Rvd24iOiJ6b25lX251bWJlciJ9XSwidGltZXNjYWxlIjoiIiwiZGlzcGxheUxlZ2VuZCI6dHJ1ZSwiYWxpZ25Nb250aCI6dHJ1ZX0%3D) .
    - Content: Spatial representation of the parking sensor data that can be used for mapping sensor locations.
    - Data type: Spatial open data.
    - Licence: CC BY / Creative Commons Attribution 4.0.
    - Limitations: Spatial points help show location, but they do not on their own explain why some areas are more congested than others.
    - Privacy/Ethics: Public infrastructure data with low privacy risk.

3.  **On-street Parking Bays**

    - Source: City of Melbourne Open Data Portal.
    - Link: (data.melbourne.vic.gov.au). [On-street Parking Bay Sensors](https://data.melbourne.vic.gov.au/explore/dataset/on-street-parking-bay-sensors/information/?dataChart=eyJxdWVyaWVzIjpbeyJjaGFydHMiOlt7InR5cGUiOiJjb2x1bW4iLCJmdW5jIjoiQ09VTlQiLCJ5QXhpcyI6ImJheV9pZCIsInNjaWVudGlmaWNEaXNwbGF5Ijp0cnVlLCJjb2xvciI6InJhbmdlLWN1c3RvbSJ9XSwieEF4aXMiOiJsYXN0dXBkYXRlZCIsIm1heHBvaW50cyI6NTAsInNvcnQiOiIiLCJ0aW1lc2NhbGUiOiJkYXkiLCJjb25maWciOnsiZGF0YXNldCI6Im9uLXN0cmVldC1wYXJraW5nLWJheS1zZW5zb3JzIiwib3B0aW9ucyI6e319LCJzZXJpZXNCcmVha2Rvd24iOiJ6b25lX251bWJlciJ9XSwidGltZXNjYWxlIjoiIiwiZGlzcGxheUxlZ2VuZCI6dHJ1ZSwiYWxpZ25Nb250aCI6dHJ1ZX0%3D) .
    - Content: Spatial polygons representing parking bays across the city.
    - Data type: Spatial administrative open data.
    - Licence/conditions: Public open data with City of Melbourne disclaimer noted on the dataset page.
    - Use in project: Optional supporting dataset for richer mapping and for understanding joins between the live parking release datasets.
    - Limitation: Not every related parking dataset joins directly to every other dataset; the sensor data joins to the parking bays data by `marker_id`.

4.  **Municipal Boundary**

    - Source: City of Melbourne / Victorian Government Data Directory.
    - Link: \[Municipal boundary\](https://data.melbourne.vic.gov.au/explore/dataset/municipal-boundary/.
    - Content: Boundary of the City of Melbourne municipality used as the basemap polygon for the spatial plots.
    - Data type: Spatial boundary data.
    - Use in project: Provides a recognisable Melbourne map so the parking points can be plotted on the municipality rather than on blank coordinate axes.

#### Data download

The following section describes how the data can be downloaded from its original sources and the initial processing steps used for analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(here)
library(lubridate)
library(sf)
library(janitor)
library(openxlsx)
library(viridis)
```

```{r}
dir.create(here::here("raw_data"), showWarnings = FALSE)
dir.create(here::here("data"), showWarnings = FALSE)
```

**Downloading Dataset 1**

```{r}
url_csv <- "https://data.melbourne.vic.gov.au/explore/dataset/on-street-parking-bay-sensors/download/?format=csv&timezone=Australia/Melbourne&lang=en"

destfile_csv <- here::here("raw_data/on-street-parking-bay-sensors.csv")
download.file(url_csv, destfile_csv, mode = "wb")

parking_sensor_raw <- read_delim(
  destfile_csv,
  delim = ";",
  show_col_types = FALSE
) |>
  clean_names()
```

If the direct download link changes, the file can be downloaded manually from the dataset page and saved as `raw_data/on-street-parking-bay-sensors.csv`.

**Downloading Dataset 2**

```{r}
url_geojson <- "https://data.melbourne.vic.gov.au/explore/dataset/on-street-parking-bay-sensors/download/?format=geojson&timezone=Australia/Melbourne&lang=en"

destfile_geojson <- here::here("raw_data/on-street-parking-bay-sensors.geojson")
download.file(url_geojson, destfile_geojson, mode = "wb")

parking_sensor_geo <- st_read(destfile_geojson, quiet = TRUE)
```

If working with the downloaded shapefile instead, all associated files (`.shp`, `.shx`, `.dbf`, `.prj`, `.cpg`) should be stored together in the same folder before reading the `.shp` file.

```{r}
# If using a shapefile instead of geojson:
# parking_sensor_geo <- st_read(here::here("raw_data/on-street-parking-bay-sensors/on-street-parking-bay-sensors.shp"))
```

**Downloading Dataset 3**

```{r}
url_bays <- "https://data.melbourne.vic.gov.au/explore/dataset/on-street-parking-bays/download/?format=csv&timezone=Australia/Melbourne&lang=en"

destfile_bays <- here::here("raw_data/on-street-parking-bays.csv")
download.file(url_bays, destfile_bays, mode = "wb")

parking_bays_raw <- read_delim(
  destfile_bays,
  delim = ";",
  show_col_types = FALSE
) |>
  clean_names()
```

This supporting dataset is useful for documenting how the City of Melbourne’s live parking datasets relate to each other.

#### Data cleaning and preparation

Before analysing congestion, the raw parking sensor data needs to be converted into a form that is easier to summarise. That includes parsing timestamps properly, extracting useful time-based variables such as hour and day of week, converting latitude and longitude into numeric form, and creating a simple logical indicator that marks whether a bay is occupied.

The main steps used to process the data are listed below.

1.  Import the raw CSV sensor data.
2.  Import the geojson or shapefile spatial data.
3.  Clean variable names.
4.  Parse timestamp fields into proper datetime format.
5.  Create derived variables such as hour, weekday, and date.
6.  Recode `status_description` into an occupied indicator.
7.  Remove or flag missing zone values.
8.  Split the coordinate field into latitude and longitude.
9.  Summarise occupancy by zone, hour, and weekday.
10. Save the cleaned analysis file separately from the raw data.

```{r}
parking_clean <- parking_sensor_raw |>
  mutate(
    lastupdated = ymd_hms(lastupdated),
    status_timestamp = ymd_hms(status_timestamp),
    hour = hour(status_timestamp),
    day_of_week = wday(status_timestamp, label = TRUE, week_start = 1),
    date = as_date(status_timestamp),
    occupied = case_when(
      str_trim(str_to_lower(status_description)) == "present" ~ TRUE,
      str_trim(str_to_lower(status_description)) == "unoccupied" ~ FALSE,
      TRUE ~ NA
    )
  ) |>
  separate(location, into = c("latitude", "longitude"), sep = ",", remove = FALSE) |>
  mutate(
    latitude = as.numeric(str_trim(latitude)),
    longitude = as.numeric(str_trim(longitude))
  ) |>
  filter(
    !is.na(zone_number),
    !is.na(status_timestamp),
    !is.na(hour),
    !is.na(day_of_week),
    !is.na(latitude),
    !is.na(longitude),
    !is.na(occupied)
  )

parking_clean

write_csv(parking_clean, here::here("data/parking_processed.csv"))
```

The cleaning step also removes rows with missing values in the key analytical fields. Based on the output shown below, the cleaned dataset contains `r nrow(parking_clean)` rows and `r ncol(parking_clean)` columns, which is the dataset used in the later summaries and visualisations

Since spatial summaries are needed, the processed summaries can later be joined to the spatial data. The City of Melbourne states that the live sensor dataset joins to the parking bays dataset by `marker_id`, and the sensor dataset joins to the parking restrictions dataset by `bay_id`.

#### Creating the data dictionary file

```{r}
data_dictionary <- tibble(
  variable = c(
    "lastupdated",
    "status_timestamp",
    "zone_number",
    "status_description",
    "kerbsideid",
    "location",
    "hour",
    "day_of_week",
    "date",
    "occupied"
  ),
  description = c(
    "Time the record was last updated",
    "Timestamp of the parking sensor reading",
    "Parking zone identifier",
    "Description of whether a car is present or not present",
    "Kerbside identifier",
    "Coordinate/location field",
    "Hour extracted from status_timestamp",
    "Weekday extracted from status_timestamp",
    "Calendar date extracted from status_timestamp",
    "Logical occupied/vacant indicator created for analysis"
  ),
  role_in_analysis = c(
    "Checks update recency",
    "Main time variable",
    "Main grouping variable",
    "Defines congestion proxy",
    "Optional finer location grouping",
    "Supports mapping",
    "Used for hourly summaries",
    "Used for weekly summaries",
    "Used for date-based summaries",
    "Main congestion measure"
  )
)

write.xlsx(data_dictionary, here::here("data/data-dictionary.xlsx"))
```

#### Files included in submission

The submission will include:

- the raw CSV file;
- the raw GeoJSON or shapefile;
- the optional raw parking bays file;
- the Melbourne municipal boundary GeoJSON;
- the processed CSV file;
- a readme file describing the source and folder structure;
- and a data dictionary saved as an `.xlsx` file.

This documentation makes it clear what was downloaded, what was changed, and what was used in the analysis.

**V1 Reflection:**

In this early version, I focused only on getting the data loaded into R. Instead of manually downloading the dtasets from the City of Melbourne website as I did on the submission for early feedback, and then loading them into R, this time I tried to download data directly from the url on Rstudio. However, this time, I noticed that when I download the datasets from the url using R markdown, the downloaded CSV files used a semicolon delimiter, so the dataset loaded as a single column. This is something I will have to fix before proceeding with any further analysis, so at this stage, the main focus is simply getting the file structure and dataset loading working.

**V2 Reflection:**

In this version, after realizing the delimiter problem, I fixed the import issue from the first draft using `read_delim(..., delim = ";")` and successfully imported the data into a format that could actually be analysed instead of being read as one long column. I then began cleaning the key variables so the timestamps, zone numbers, and occupancy values could be used properly and also started creating the time-based fields like hour and day of week.

Another thing I learned at this stage was that the location field needed extra wrangling before it could be used in maps. Splitting coordinates into latitude and longitude and checking for missing values took a bit of trial and error, but it was necessary before I could move on to any spatial visualisation. This version still feels very preliminary, but is a major step forward because the dataset is now structured enough to support summaries by time and place and I've also added a data dictionary file to supplement understanding.
